# Classification

## Conceptual

### Question 1

> Using a little bit of algebra, prove that (4.2) is equivalent to (4.3). In
> other words, the logistic function representation and logit representation
> for the logistic regression model are equivalent.

We need to show that

$$
p(X) = \frac{e^{\beta_0 + \beta_1X}}{1 + e^{\beta_0 + \beta_1X}}
$$

is equivalent to

$$
\frac{p(X)}{1-p(X)} = e^{\beta_0 + \beta_1X}
$$

Letting $x = e^{\beta_0 + \beta_1X}$

\begin{align}
\frac{P(X)}{1-p(X)} &= \frac{\frac{x}{1 + x}}
                            {1 - \frac{x}{1 + x}} \\
              &= \frac{\frac{x}{1 + x}}
                      {\frac{1}{1 + x}} \\
              &= x
\end{align}

### Question 2

> It was stated in the text that classifying an observation to the class for
> which (4.12) is largest is equivalent to classifying an observation to the
> class for which (4.13) is largest. Prove that this is the case. In other
> words, under the assumption that the observations in the $k$th class are
> drawn from a $N(\mu_k,\sigma^2)$ distribution, the Bayes' classifier assigns
> an observation to the class for which the discriminant function is maximized.

4.12 is

$$
p_k(x) = \frac{\pi_k\frac{1}{\sqrt{2\pi\sigma}} \exp(-\frac{1}{2\sigma^2}(x - \mu_k)^2)}
              {\sum_{l=1}^k \pi_l\frac{1}{\sqrt{2\pi\sigma}} \exp(-\frac{1}{2\sigma^2}(x - \mu_l)^2)}
$$

and the discriminant function is

$$
\delta_k(x) = x.\frac{\mu_k}{\sigma^2} - \frac{\mu_k^2}{2\sigma_2} + \log(\pi_k)
$$


Since $\sigma^2$ is constant

$$
p_k(x) = \frac{\pi_k \exp\left(-\frac{1}{2\sigma^2}(x - \mu_k)^2\right)}
              {\sum_{l=1}^k \pi_l \exp\left(-\frac{1}{2\sigma^2}(x - \mu_l)^2\right)}
$$

Maximizing $p_k(x)$ also maximizes any monotonic function of $p_k(X)$, and
therefore, we can consider maximizing $\log(p_K(X))$

$$
\log(p_k(x)) = \log(\pi_k) - \frac{1}{2\sigma^2}(x - \mu_k)^2 -
              \log\left(\sum_{l=1}^k \pi_l \exp\left(-\frac{1}{2\sigma^2}(x - \mu_l)^2\right)\right)
$$

Remember that we are maximizing over $k$, and since the last term does not
vary with $k$ it can be ignored. So we just need to maximize

\begin{align}
f &= \log(\pi_k) - \frac{1}{2\sigma^2} (x^2 - 2x\mu_k + \mu_k^2) \\
  &= \log(\pi_k) - \frac{x^2}{2\sigma^2} + \frac{x\mu_k}{\sigma^2} - \frac{\mu_k^2}{2\sigma^2} \\
\end{align}

Since $\frac{x^2}{2\sigma^2}$ is also independent of $k$, we just need to
maximize

$$
\log(\pi_k) + \frac{x\mu_k}{\sigma^2} - \frac{\mu_k^2}{2\sigma^2}
$$

### Question 3

> This problem relates to the QDA model, in which the observations within each
> class are drawn from a normal distribution with a class-specific mean vector
> and a class specific covariance matrix. We consider the simple case where $p
> = 1$; i.e. there is only one feature.
>
> Suppose that we have $K$ classes, and that if an observation belongs to the
> $k$th class then $X$ comes from a one-dimensional normal distribution,
> $X \sim N(\mu_k,\sigma^2)$. Recall that the density function for the
> one-dimensional normal distribution is given in (4.16). Prove that in this
> case, the Bayes classifier is _not_ linear. Argue that it is in fact
> quadratic.
>
> _Hint: For this problem, you should follow the arguments laid out in_
> _Section 4.4.1, but without making the assumption that_
> _$\sigma_1^2 = ... = \sigma_K^2$._

As above,

$$
p_k(x) = \frac{\pi_k\frac{1}{\sqrt{2\pi\sigma_k}} \exp(-\frac{1}{2\sigma_k^2}(x - \mu_k)^2)}
              {\sum_{l=1}^k \pi_l\frac{1}{\sqrt{2\pi\sigma_l}} \exp(-\frac{1}{2\sigma_l^2}(x - \mu_l)^2)}
$$

Now lets derive the Bayes classifier, without assuming 
$\sigma_1^2 = ... = \sigma_K^2$

Maximizing $p_k(x)$ also maximizes any monotonic function of $p_k(X)$, and
therefore, we can consider maximizing $\log(p_K(X))$

$$
\log(p_k(x)) = \log(\pi_k) + \log\left(\frac{1}{\sqrt{2\pi\sigma_k}}\right) - \frac{1}{2\sigma_k^2}(x - \mu_k)^2 -
              \log\left(\sum_{l=1}^k \frac{1}{\sqrt{2\pi\sigma_l}} \pi_l \exp\left(-\frac{1}{2\sigma_l^2}(x - \mu_l)^2\right)\right)
$$

Remember that we are maximizing over $k$, and since the last term does not
vary with $k$ it can be ignored. So we just need to maximize

\begin{align}
f &= \log(\pi_k) + \log\left(\frac{1}{\sqrt{2\pi\sigma_k}}\right) - \frac{1}{2\sigma_k^2}(x - \mu_k)^2 \\
  &= \log(\pi_k) + \log\left(\frac{1}{\sqrt{2\pi\sigma_k}}\right) - \frac{x^2}{2\sigma_k^2} + \frac{x\mu_k}{\sigma_k^2} - \frac{\mu_k^2}{2\sigma_k^2}  \\
\end{align}

However, unlike in Q2, $\frac{x^2}{2\sigma_k^2}$ is not independent of $k$, so
we retain the term with $x^2$, hence $f$, the Bayes’ classifier, is a
quadratic function of $x$.

### Question 4

> When the number of features $p$ is large, there tends to be a deterioration
> in the performance of KNN and other _local_ approaches that perform
> prediction using only observations that are _near_ the test observation for
> which a prediction must be made. This phenomenon is known as the
> _curse of dimensionality_, and it ties into the fact that non-parametric
> approaches often perform poorly when $p$ is large. We will now investigate
> this curse.
>
> a. Suppose that we have a set of observations, each with measurements on
>    $p = 1$ feature, $X$. We assume that $X$ is uniformly (evenly) distributed
>    on $[0, 1]$. Associated with each observation is a response value. Suppose
>    that we wish to predict a test observation's response using only
>    observations that are within 10% of the range of $X$ closest to that test
>    observation. For instance, in order to predict the response for a test
>    observation with $X = 0.6$, we will use observations in the range
>    $[0.55, 0.65]$. On average, what fraction of the available observations
>    will we use to make the prediction?

For values in $0...0.05$, we use less than 10% of observations (between 5% and
10%, 7.5% on average), similarly with values in $0.95...1$. For values in
$0.05...0.95$ we use 10% of available observations. The (weighted) average is
then $7.5 \times 0.1 + 10 \times 0.9 = 9.75\%$.

> b. Now suppose that we have a set of observations, each with measurements on
>    $p = 2$ features, $X_1$ and $X_2$. We assume that $(X_1, X_2)$ are
>    uniformly distributed on $[0, 1] \times [0, 1]$. We wish to predict a test
>    observation's response using only observations that are within 10% of the
>    range of $X_1$ _and_ within 10% of the range of $X_2$ closest to that test
>    observation. For instance, in order to predict the response for a test
>    observation with $X_1 = 0.6$ and $X_2 = 0.35$, we will use observations in
>    the range $[0.55, 0.65]$ for $X_1$ and in the range $[0.3, 0.4]$ for $X_2$.
>    On average, what fraction of the available observations will we use to
>    make the prediction?

Since we need the observation to be within range for $X_1$ and $X_2$ we square
9.75% = $0.0975^2 \times 100 = 0.95\%$

> c. Now suppose that we have a set of observations on $p = 100$ features. Again
>    the observations are uniformly distributed on each feature, and again each
>    feature ranges in value from 0 to 1. We wish to predict a test
>    observation's response using observations within the 10% of each feature's
>    range that is closest to that test observation. What fraction of the
>    available observations will we use to make the prediction?

Similar to above, we use: $0.0975^{100} \times 100 = 8 \times 10^{-100}\%$,
essentially zero.

> d. Using your answers to parts (a)--(c), argue that a drawback of KNN when
>    $p$ is large is that there are very few training observations "near" any
>    given test observation.

As $p$ increases, the fraction of observations near any given point rapidly
approaches zero. For instance, even if you use 50% of the nearest observations
for each $p$, with $p = 10$, only $0.5^{10} \times 100 \approx 0.1\%$ points are
"near".

> e. Now suppose that we wish to make a prediction for a test observation by
>    creating a $p$-dimensional hypercube centered around the test observation
>    that contains, on average, 10% of the training observations. For
>    $p = 1,2,$ and $100$, what is the length of each side of the hypercube?
>    Comment on your answer.
>
> _Note: A hypercube is a generalization of a cube to an arbitrary number of_
> _dimensions. When $p = 1$, a hypercube is simply a line segment, when $p = 2$_
> _it is a square, and when $p = 100$ it is a 100-dimensional cube._

When $p = 1$, clearly the length is 0.1.
When $p = 2$, we need the value $l$ such that $l^2 = 0.1$, so
$l = \sqrt{0.1} = 0.32$.
When $p = n$, $l = 0.1^{1/n}$, so in the case of $n = 100$, $l = 0.98$.
Therefore, the length of each side of the hypercube rapidly approaches 1
(or 100%) of the range of each $p$.

### Question 5

> We now examine the differences between LDA and QDA.
>
> a. If the Bayes decision boundary is linear, do we expect LDA or QDA to
>    perform better on the training set? On the test set?

QDA, being a more flexible model, will always perform better on the training
set, but LDA would be expected to perform better on the test set.

> b. If the Bayes decision boundary is non-linear, do we expect LDA or QDA to
>    perform better on the training set? On the test set?

QDA, being a more flexible model, will perform better on the training
set, and we would hope that extra flexibility translates to a better fit on
the test set.

> c. In general, as the sample size $n$ increases, do we expect the test
>    prediction accuracy of QDA relative to LDA to improve, decline, or be
>    unchanged? Why?

As $n$ increases, we would expect the prediction accuracy of QDA relative to
LDA to improve as there is more data to fit to subtle effects in the data.

> d. True or False: Even if the Bayes decision boundary for a given problem is
>    linear, we will probably achieve a superior test error rate using QDA
>    rather than LDA because QDA is flexible enough to model a linear decision
>    boundary. Justify your answer.

False. QDA can overfit leading to poorer test performance.

### Question 6

> Suppose we collect data for a group of students in a statistics class with
> variables $X_1 =$ hours studied, $X_2 =$ undergrad GPA, and $Y =$ receive an A.
> We fit a logistic regression and produce estimated coefficient,
> $\hat\beta_0 = -6$, $\hat\beta_1 = 0.05$, $\hat\beta_2 = 1$.
>
> a. Estimate the probability that a student who studies for 40h and has an
>    undergrad GPA of 3.5 gets an A in the class.

The logistic model is:

$$
\log\left(\frac{p(X)}{1-p(x)}\right) = -6 + 0.05X_1 + X_2
$$

or

$$
p(X) = \frac{e^{-6 + 0.05X_1 + X_2}}{1 + e^{-6 + 0.05X_1 + X_2}}
$$

when $X_1 = 40$ and $X_2 = 3.5$, $p(X) = 0.38$

> b. How many hours would the student in part (a) need to study to have a 50%
>    chance of getting an A in the class?

We would like to solve for $X_1$ where $p(X) = 0.5$. Taking the first equation
above, we need to solve $0 = −6 + 0.05X_1 + 3.5$, so $X_1 = 50$ hours.

### Question 7

> Suppose that we wish to predict whether a given stock will issue a dividend
> this year ("Yes" or "No") based on $X$, last year's percent profit. We
> examine a large number of companies and discover that the mean value of $X$
> for companies that issued a dividend was $\bar{X} = 10$, while the mean for
> those that didn't was $\bar{X} = 0$. In addition, the variance of $X$ for
> these two sets of companies was $\hat{\sigma}^2 = 36$. Finally, 80% of
> companies issued dividends. Assuming that $X$ follows a normal distribution,
> predict the probability that a company will issue a dividend this year given
> that its percentage profit was $X = 4$ last year.
>
> _Hint: Recall that the density function for a normal random variable is_
> _$f(x) =\frac{1}{\sqrt{2\pi\sigma^2}}e^{-(x-\mu)^2/2\sigma^2}$._
> _You will need to use Bayes' theorem._

Value $v$ for companies (D) issuing a dividend = $v_D \sim \mathcal{N}(10, 36)$.
Value $v$ for companies (N) not issuing a dividend = $v_N \sim \mathcal{N}(0, 36)$
and $p(D) = 0.8$.

We want to find $p(D|v)$ and we can calculate $p(v|D)$ from the Gaussian
density function. Note that since $e^2$ is equal between both classes, the
term $\frac{1}{\sqrt{2\pi\sigma^2}}$ cancels.

\begin{align}
p(D|v) &= \frac{p(v|D) p(D)}{p(v|D)p(D) + p(v|N)p(N)} \\
       &= \frac{\pi_D \frac{1}{\sqrt{2\pi\sigma^2}} e^{-(x-\mu_D)^2/2\sigma^2}}
               {\pi_D \frac{1}{\sqrt{2\pi\sigma^2}} e^{-(x-\mu_D)^2/2\sigma^2} +
                \pi_N \frac{1}{\sqrt{2\pi\sigma^2}} e^{-(x-\mu_N)^2/2\sigma^2}} \\
       &= \frac{\pi_D e^{-(x-\mu_D)^2/2\sigma^2}}
               {\pi_D e^{-(x-\mu_D)^2/2\sigma^2} +
                \pi_N e^{-(x-\mu_N)^2/2\sigma^2}} \\
       &= \frac{0.8 \times e^{-(4-10)^2/(2 \times 36)}}
               {0.8 \times e^{-(4-10)^2/(2 \times 36)} + 0.2 \times e^{-(4-0)^2/(2 \times 36)}} \\
       &= \frac{0.8 e^{-1/2}}{0.8 e^{-1/2} + 0.2 e^{-2/9}}
\end{align}

```{r}
exp(-0.5) * 0.8 / (exp(-0.5) * 0.8 + exp(-2/9) * 0.2)
```

### Question 8

> Suppose that we take a data set, divide it into equally-sized training and
> test sets, and then try out two different classification procedures. First we
> use logistic regression and get an error rate of 20% on the training data and
> 30% on the test data. Next we use 1-nearest neighbors (i.e. $K = 1$) and get
> an average error rate (averaged over both test and training data sets) of
> 18%. Based on these results, which method should we prefer to use for
> classification of new observations? Why?

For $K = 1$, performance on the training set is perfect and the error rate
is zero, implying a test error rate of 36%. Logistic regression outperforms
1-nearest neighbor on the test set and therefore should be preferred.

### Question 9

> This problem has to do with _odds_.
>
> a. On average, what fraction of people with an odds of 0.37 of defaulting on
>    their credit card payment will in fact default?

Odds is defined as $p/(1-p)$.

$$0.37 = \frac{p(x)}{1 - p(x)}$$

therefore,

$$p(x) = \frac{0.37}{1 + 0.37} = 0.27$$

> b. Suppose that an individual has a 16% chance of defaulting on her credit
>    card payment. What are the odds that she will default?

$$0.16 / (1 - 0.16)  = 0.19$$

### Question 10

> Equation 4.32 derived an expression for $\log(\frac{Pr(Y=k|X=x)}{Pr(Y=K|X=x)})$
> in the setting where $p > 1$, so that the mean for the $k$th class, $\mu_k$,
> is a $p$-dimensional vector, and the shared covariance $\Sigma$ is a 
> $p \times p$ matrix. However, in the setting with $p = 1$, (4.32) takes a 
> simpler form, since the means $\mu_1, ..., \mu_k$ and the variance $\sigma^2$
> are scalars. In this simpler setting, repeat the calculation in (4.32), and 
> provide expressions for $a_k$ and $b_{kj}$ in terms of 
> $\pi_k, \pi_K, \mu_k, \mu_K,$ and $\sigma^2$.

\begin{align*}
\log\left(\frac{Pr(Y=k|X=x)}{Pr(Y=K|X=x)}\right)
 & = \log\left(\frac{\pi_k f_k(x)}{\pi_K f_K(x)}\right) \\
 & = \log\left(\frac{\pi_k \exp(-1/2((x-\mu_k)/\sigma)^2)}{\pi_K \exp(-1/2((x-\mu_K)/\sigma)^2)}\right) \\
 & = \log\left(\frac{\pi_k}{\pi_K}\right) - \frac{1}{2} \left(\frac{x-\mu_k}{\sigma}\right)^2 + \frac{1}{2} \left(\frac{x-\mu_K}{\sigma}\right)^2 \\
 & = \log\left(\frac{\pi_k}{\pi_K}\right) - \frac{1}{2\sigma^2} (x-\mu_k)^2 + \frac{1}{2\sigma^2} (x-\mu_K)^2 \\
 & = \log\left(\frac{\pi_k}{\pi_K}\right) - \frac{1}{2\sigma^2} \left((x-\mu_k)^2 - (x-\mu_K)^2\right) \\
 & = \log\left(\frac{\pi_k}{\pi_K}\right) - \frac{1}{2\sigma^2} \left(x^2-2x\mu_k+\mu_k^2 - x^2 + 2x\mu_K - \mu_K^2\right) \\
 & = \log\left(\frac{\pi_k}{\pi_K}\right) - \frac{1}{2\sigma^2} \left(2x(\mu_K - \mu_k) + \mu_k^2 -\mu_K^2\right) \\
 & = \log\left(\frac{\pi_k}{\pi_K}\right) - \frac{\mu_k^2 -\mu_K^2}{2\sigma^2} + \frac{x(\mu_k - \mu_K)}{\sigma^2}
\end{align*}

Therefore,

$$a_k = \log\left(\frac{\pi_k}{\pi_K}\right) - \frac{\mu_k^2 -\mu_K^2}{2\sigma^2}$$

and

$$b_k = (\mu_k - \mu_K) / \sigma^2$$

### Question 11

ToDo

> Work out the detailed forms of $a_k$, $b_{kj}$, and $b_{kjl}$ in (4.33).
> Your answer should involve $\pi_k$, $\pi_K$, $\mu_k$, $\mu_K$, $\Sigma_k$, 
> and $\Sigma_K$.

### Question 12

> Suppose that you wish to classify an observation $X \in \mathbb{R}$ into 
> `apples` and `oranges`. You fit a logistic regression model and find that
>
> $$
> \hat{Pr}(Y=orange|X=x) = 
> \frac{\exp(\hat\beta_0 + \hat\beta_1x)}{1 + \exp(\hat\beta_0 + \hat\beta_1x)}
> $$
>
> Your friend fits a logistic regression model to the same data using the
> _softmax_ formulation in (4.13), and finds that
>
> $$
> \hat{Pr}(Y=orange|X=x) = 
> \frac{\exp(\hat\alpha_{orange0} + \hat\alpha_{orange1}x)}
> {\exp(\hat\alpha_{orange0} + \hat\alpha_{orange1}x) + \exp(\hat\alpha_{apple0} + \hat\alpha_{apple1}x)}
> $$
> 
> a. What is the log odds of `orange` versus `apple` in your model?

The log odds is just $\hat\beta_0 + \hat\beta_1x$

> b. What is the log odds of `orange` versus `apple` in your friend's model?

From 4.14, log odds of our friend's model is:

$$
(\hat\alpha_{orange0} - \hat\alpha_{apple0}) + (\hat\alpha_{orange1} - \hat\alpha_{apple1})x
$$

> c. Suppose that in your model, $\hat\beta_0 = 2$ and $\hat\beta = −1$. What
>    are the coefficient estimates in your friend's model? Be as specific as
>    possible.

We can say that in our friend's model $\hat\alpha_{orange0} -
\hat\alpha_{apple0} = 2$ and $\hat\alpha_{orange1} - \hat\alpha_{apple1} = -1$.

We are unable to know the specific value of each parameter however.

> d. Now suppose that you and your friend fit the same two models on a different
>    data set. This time, your friend gets the coefficient estimates 
>    $\hat\alpha_{orange0} = 1.2$, $\hat\alpha_{orange1} = −2$, 
>    $\hat\alpha_{apple0} = 3$, $\hat\alpha_{apple1} = 0.6$. What are the 
>    coefficient estimates in your model?

The coefficients in our model would be $\hat\beta_0 = 1.2 - 3 = -1.8$ and
$\hat\beta_1 = -2 - 0.6 = -2.6$

> e. Finally, suppose you apply both models from (d) to a data set with 2,000
>    test observations. What fraction of the time do you expect the predicted 
>    class labels from your model to agree with those from your friend's model?
>    Explain your answer.

The models are identical with different parameterization so they should 
perfectly agree.

## Applied

### Question 13

> This question should be answered using the `Weekly` data set, which is part
> of the `ISLR2` package. This data is similar in nature to the `Smarket` data
> from this chapter's lab, except that it contains 1,089 weekly returns for 21
> years, from the beginning of 1990 to the end of 2010.
>
> a. Produce some numerical and graphical summaries of the `Weekly` data. Do
>    there appear to be any patterns?

```{r, message = FALSE, warning = FALSE}
library(MASS)
library(class)
library(tidyverse)
library(corrplot)
library(ISLR2)
library(e1071)
```

```{r}
summary(Weekly)
corrplot(cor(Weekly[, -9]), type = "lower", diag = FALSE, method = "ellipse")
```

Volume is strongly positively correlated with Year. Other correlations are
week, but Lag1 is negatively correlated with Lag2 but positively correlated
with Lag3.

> b. Use the full data set to perform a logistic regression with `Direction` as
>    the response and the five lag variables plus `Volume` as predictors. Use
>    the summary function to print the results. Do any of the predictors appear
>    to be statistically significant? If so, which ones?

```{r}
fit <- glm(
  Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume,
  data = Weekly,
  family = binomial
)
summary(fit)
```

Lag2 is significant.

> c. Compute the confusion matrix and overall fraction of correct predictions.
>    Explain what the confusion matrix is telling you about the types of
>    mistakes made by logistic regression.

```{r}
contrasts(Weekly$Direction)
pred <- predict(fit, type = "response") > 0.5
(t <- table(ifelse(pred, "Up (pred)", "Down (pred)"), Weekly$Direction))
sum(diag(t)) / sum(t)
```

The overall fraction of correct predictions is 0.56. Although logistic
regression correctly predicts upwards movements well, it incorrectly predicts
most downwards movements as up.

> d. Now fit the logistic regression model using a training data period from
>    1990 to 2008, with `Lag2` as the only predictor. Compute the confusion
>    matrix and the overall fraction of correct predictions for the held out
>    data (that is, the data from 2009 and 2010).

```{r}
train <- Weekly$Year < 2009

fit <- glm(Direction ~ Lag2, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
(t <- table(ifelse(pred, "Up (pred)", "Down (pred)"), Weekly[!train, ]$Direction))
sum(diag(t)) / sum(t)
```

> e. Repeat (d) using LDA.

```{r}
fit <- lda(Direction ~ Lag2, data = Weekly[train, ])
pred <- predict(fit, Weekly[!train, ], type = "response")$class
(t <- table(pred, Weekly[!train, ]$Direction))
sum(diag(t)) / sum(t)
```

> f. Repeat (d) using QDA.

```{r}
fit <- qda(Direction ~ Lag2, data = Weekly[train, ])
pred <- predict(fit, Weekly[!train, ], type = "response")$class
(t <- table(pred, Weekly[!train, ]$Direction))
sum(diag(t)) / sum(t)
```

> g. Repeat (d) using KNN with $K = 1$.

```{r}
fit <- knn(
  Weekly[train, "Lag2", drop = FALSE],
  Weekly[!train, "Lag2", drop = FALSE],
  Weekly$Direction[train]
)
(t <- table(fit, Weekly[!train, ]$Direction))
sum(diag(t)) / sum(t)
```

> h. Repeat (d) using naive Bayes.

```{r}
fit <- naiveBayes(Direction ~ Lag2, data = Smarket, subset = train)
pred <- predict(fit, Weekly[!train, ], type = "class")
(t <- table(pred, Weekly[!train, ]$Direction))
sum(diag(t)) / sum(t)
```

> i. Which of these methods appears to provide the best results on this data?

Logistic regression and LDA are the best performing.

> j. Experiment with different combinations of predictors, including possible
>    transformations and interactions, for each of the methods. Report the
>    variables, method, and associated confusion matrix that appears to provide
>    the best results on the held out data. Note that you should also
>    experiment with values for $K$ in the KNN classifier.

```{r}
fit <- glm(Direction ~ Lag1, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
mean(ifelse(pred, "Up", "Down") == Weekly[!train, ]$Direction)

fit <- glm(Direction ~ Lag3, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
mean(ifelse(pred, "Up", "Down") == Weekly[!train, ]$Direction)

fit <- glm(Direction ~Lag4, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
mean(ifelse(pred, "Up", "Down") == Weekly[!train, ]$Direction)

fit <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
mean(ifelse(pred, "Up", "Down") == Weekly[!train, ]$Direction)

fit <- glm(Direction ~ Lag1 * Lag2 * Lag3 * Lag4, data = Weekly[train, ], family = binomial)
pred <- predict(fit, Weekly[!train, ], type = "response") > 0.5
mean(ifelse(pred, "Up", "Down") == Weekly[!train, ]$Direction)

fit <- lda(Direction ~ Lag1 + Lag2 + Lag3 + Lag4,data = Weekly[train, ])
pred <- predict(fit, Weekly[!train, ], type = "response")$class
mean(pred == Weekly[!train, ]$Direction)

fit <- qda(Direction ~ Lag1 + Lag2 + Lag3 + Lag4, data = Weekly[train, ])
pred <- predict(fit, Weekly[!train, ], type = "response")$class
mean(pred == Weekly[!train, ]$Direction)

fit <- naiveBayes(Direction ~ Lag1 + Lag2 + Lag3 + Lag4, data = Weekly[train, ])
pred <- predict(fit, Weekly[!train, ], type = "class")
mean(pred == Weekly[!train, ]$Direction)

set.seed(1)
res <- sapply(1:30, function(k) {
  fit <- knn(
    Weekly[train, 2:4, drop = FALSE],
    Weekly[!train, 2:4, drop = FALSE],
    Weekly$Direction[train],
    k = k
  )
  mean(fit == Weekly[!train, ]$Direction)
})
plot(1:30, res, type = "o", xlab = "k", ylab = "Fraction correct")
(k <- which.max(res))

fit <- knn(
  Weekly[train, 2:4, drop = FALSE],
  Weekly[!train, 2:4, drop = FALSE],
  Weekly$Direction[train],
  k = k
)
table(fit, Weekly[!train, ]$Direction)
mean(fit == Weekly[!train, ]$Direction)
```

KNN using the first 3 Lag variables performs marginally better than logistic
regression with `Lag2` if we tune $k$ to be $k = 26$.

### Question 14

> In this problem, you will develop a model to predict whether a given car gets
> high or low gas mileage based on the `Auto` data set.
>
> a. Create a binary variable, `mpg01`, that contains a 1 if `mpg` contains a
>    value above its median, and a 0 if `mpg` contains a value below its
>    median. You can compute the median using the `median()` function. Note you
>    may find it helpful to use the `data.frame()` function to create a single
>    data set containing both `mpg01` and the other `Auto` variables.

```{r}
x <- cbind(Auto[, -1], data.frame("mpg01" = Auto$mpg > median(Auto$mpg)))
```

> b. Explore the data graphically in order to investigate the association
>    between `mpg01` and the other features. Which of the other features seem
>    most likely to be useful in predicting `mpg01`? Scatterplots and boxplots
>    may be useful tools to answer this question. Describe your findings.

```{r}
par(mfrow = c(2, 4))
for (i in 1:7) hist(x[, i], breaks = 20, main = colnames(x)[i])

par(mfrow = c(2, 4))
for (i in 1:7) boxplot(x[, i] ~ x$mpg01, main = colnames(x)[i])

pairs(x[, 1:7])
```

Most variables show an association with `mpg01` category, and several
variables are colinear.

> c. Split the data into a training set and a test set.

```{r}
set.seed(1)
train <- sample(seq_len(nrow(x)), nrow(x) * 2/3)
```

> d. Perform LDA on the training data in order to predict `mpg01` using the
>    variables that seemed most associated with `mpg01` in (b). What is the
>    test error of the model obtained?

```{r}
sort(sapply(1:7, function(i) {
  setNames(abs(t.test(x[, i] ~ x$mpg01)$statistic), colnames(x)[i])
}))

fit <- lda(mpg01 ~ cylinders + weight + displacement, data = x[train, ])
pred <- predict(fit, x[-train, ], type = "response")$class
mean(pred != x[-train, ]$mpg01)
```

> e. Perform QDA on the training data in order to predict `mpg01` using the
>    variables that seemed most associated with `mpg01` in (b). What is the
>    test error of the model obtained?

```{r}
fit <- qda(mpg01 ~ cylinders + weight + displacement, data = x[train, ])
pred <- predict(fit, x[-train, ], type = "response")$class
mean(pred != x[-train, ]$mpg01)
```

> f. Perform logistic regression on the training data in order to predict
>    `mpg01` using the variables that seemed most associated with `mpg01` in
>    (b). What is the test error of the model obtained?

```{r}
fit <- glm(mpg01 ~ cylinders + weight + displacement, data = x[train, ], family = binomial)
pred <- predict(fit, x[-train, ], type = "response") > 0.5
mean(pred != x[-train, ]$mpg01)
```

> g. Perform naive Bayes on the training data in order to predict `mpg01` using
>    the variables that seemed most associated with `mpg01` in (b). What is the
>    test error of the model obtained?

```{r}
fit <- naiveBayes(mpg01 ~ cylinders + weight + displacement, data = x[train, ])
pred <- predict(fit, x[-train, ], type = "class")
mean(pred != x[-train, ]$mpg01)
```

> h. Perform KNN on the training data, with several values of $K$, in order to
>    predict `mpg01`. Use only the variables that seemed most associated with
>    `mpg01` in (b). What test errors do you obtain? Which value of $K$ seems
>    to perform the best on this data set?

```{r}
res <- sapply(1:50, function(k) {
  fit <- knn(x[train, c(1, 4, 2)], x[-train, c(1, 4, 2)], x$mpg01[train], k = k)
  mean(fit != x[-train, ]$mpg01)
})
names(res) <- 1:50
plot(res, type = "o")
res[which.min(res)]
```

For the models tested here, $k = 32$ appears to perform best. QDA has a lower
error rate overall, performing slightly better than LDA.

### Question 15

> This problem involves writing functions.
>
> a. Write a function, `Power()`, that prints out the result of raising 2 to
>    the 3rd power. In other words, your function should compute $2^3$ and
>    print out the results.
>
>    _Hint: Recall that `x^a` raises `x` to the power `a`. Use the `print()`_
>    _function to output the result._

```{r}
Power <- function() print(2^3)
```

> b. Create a new function, `Power2()`, that allows you to pass any two
>    numbers, `x` and `a`, and prints out the value of `x^a`. You can do this
>    by beginning your function with the line
>    ```r
>    > Power2=function(x,a) {
>    ```
>    You should be able to call your function by entering, for instance,
>    ```r
>    > Power2(3, 8)
>    ```
>    on the command line. This should output the value of $3^8$, namely, 6,561.

```{r}
Power2 <- function(x, a) print(x^a)
```

> c. Using the `Power2()` function that you just wrote, compute $10^3$,
>    $8^{17}$, and $131^3$.

```{r}
c(Power2(10, 3), Power2(8, 17), Power2(131, 3))
```

> d. Now create a new function, `Power3()`, that actually returns the result
>    `x^a` as an `R` object, rather than simply printing it to the screen. That
>    is, if you store the value `x^a` in an object called result within your
>    function, then you can simply `return()` this result, using the following
>    line:
>    ```r
>    > return(result)
>    ```
>    The line above should be the last line in your function, before the `}`
>    symbol.

```{r}
Power3 <- function(x, a) {
  result <- x^a
  return(result)
}
```

> e. Now using the `Power3()` function, create a plot of $f(x) = x^2$. The
>    $x$-axis should display a range of integers from 1 to 10, and the $y$-axis
>    should display $x^2$. Label the axes appropriately, and use an appropriate
>    title for the figure. Consider displaying either the $x$-axis, the
>    $y$-axis, or both on the log-scale. You can do this by using `log = "x"`,
>    `log = "y"`, or `log = "xy"` as arguments to the `plot()` function.

```{r}
plot(1:10, Power3(1:10, 2), 
  xlab = "x", 
  ylab = expression(paste("x"^"2")),
  log = "y"
)
```

> f. Create a function, `PlotPower()`, that allows you to create a plot of `x`
>    against `x^a` for a fixed `a` and for a range of values of `x`. For
>    instance, if you call
>    ```r
>    > PlotPower(1:10, 3)
>    ```
>    then a plot should be created with an $x$-axis taking on values 
>    $1,2,...,10$, and a $y$-axis taking on values $1^3,2^3,...,10^3$.

```{r}
PlotPower <- function(x, a, log = "y") {
  plot(x, Power3(x, a),
    xlab = "x", 
    ylab = substitute("x"^a, list(a = a)),
    log = log
  )
}

PlotPower(1:10, 3)
```

### Question 13

> Using the `Boston` data set, fit classification models in order to predict
> whether a given census tract has a crime rate above or below the median.
> Explore logistic regression, LDA, naive Bayes and KNN models using various
> sub-sets of the predictors. Describe your findings.
>
> _Hint: You will have to create the response variable yourself, using the_
> _variables that are contained in the `Boston` data set._

```{r}
x <- cbind(
  ISLR2::Boston[, -1], 
  data.frame("highcrim" = Boston$crim > median(Boston$crim))
)
set.seed(1)
train <- sample(seq_len(nrow(x)), nrow(x) * 2/3)
```

We can find the most associated variables by performing wilcox tests.

```{r}
ord <- order(sapply(1:12, function(i) {
  p <- wilcox.test(as.numeric(x[train, i]) ~ x[train, ]$highcrim)$p.value
  setNames(log10(p), colnames(x)[i])
}))
ord <- names(x)[ord]
ord
```

Variables `nox` (nitrogen oxides concentration) followed by `dis` (distance to
employment center) appear to be most associated with high crime.

Let's reorder columns by those most associated with highcrim (in the training
data)

```{r}
x <- x[, c(ord, "highcrim")]
```

Let's look at univariate associations with `highcrim` (in the training data)

```{r}
x[train, ] |>
  pivot_longer(!highcrim) |>
  mutate(name = factor(name, levels = ord)) |>
  ggplot(aes(highcrim, value)) + 
  geom_boxplot() + 
  facet_wrap(~name, scale = "free")
```

Fit lda, logistic regression, naive Bayes and KNN models (with k = 1..50) for a
set of specific predictors and return the error rate. We fit models using
increasing numbers of predictors: column 1, then columns 1 and 2 etc.

```{r}
fit_models <- function(cols, k_vals = 1:50) {
  dat_train <- x[train, cols, drop = FALSE]
  dat_test <- x[-train, cols, drop = FALSE]

  fit <- lda(x$highcrim[train] ~ ., data = dat_train)
  pred <- predict(fit, dat_test, type = "response")$class
  lda_err <- mean(pred != x$highcrim[-train])

  fit <- glm(x$highcrim[train] ~ ., data = dat_train, family = binomial)
  pred <- predict(fit, dat_test, type = "response") > 0.5
  logreg_err <- mean(pred != x$highcrim[-train])

  fit <- naiveBayes(x$highcrim[train] ~ ., data = dat_train)
  pred <- predict(fit, dat_test, type = "class")
  nb_err <- mean(pred != x$highcrim[-train])

  res <- sapply(k_vals, function(k) {
    fit <- knn(dat_train, dat_test, x$highcrim[train], k = k)
    mean(fit != x$highcrim[-train])
  })
  knn_err <- min(res)

  c("LDA" = lda_err, "LR" = logreg_err, "NB" = nb_err, "KNN" = knn_err)
}
```

```{r}
res <- sapply(1:12, function(max) fit_models(1:max))
res <- as_tibble(t(res))
res$n_var <- 1:12
pivot_longer(res, cols = !n_var) |>
  ggplot(aes(n_var, value, col = name)) + 
  geom_line() + 
  xlab("Number of predictors") + 
  ylab("Error rate")
```

KNN appears to perform better (if we tune $k$) for all numbers of predictors.

```{r}
fit <- knn(
  x[train, "nox", drop = FALSE],
  x[-train, "nox", drop = FALSE],
  x$highcrim[train],
  k = 1
)
table(fit, x[-train, ]$highcrim)
mean(fit != x[-train, ]$highcrim) * 100
```

Surprisingly, the best model (with an error rate of <5%) uses $k = 1$ and
assigns crime rate categories based on the town with the single most similar
nitrogen oxide concentration (`nox`). This might be, for example, because nearby
towns have similar crime rates, and we can obtain good predictions by predicting
crime rate based on a nearby town.

But what if we only consider $k = 20$.

```{r}
res <- sapply(1:12, function(max) fit_models(1:max, k_vals = 20))
res <- as_tibble(t(res))
res$n_var <- 1:12
pivot_longer(res, cols = !n_var) |>
  ggplot(aes(n_var, value, col = name)) +
  geom_line() +
  xlab("Number of predictors") +
  ylab("Error rate")
```

KNN still performs best with a single predictor (`nox`), but logistic regression
with 12 predictors also performs well and has an error rate of ~12%.

```{r}
vars <- names(x)[1:12]
dat_train <- x[train, vars]
dat_test <- x[-train, vars]

fit <- glm(x$highcrim[train] ~ ., data = dat_train, family = binomial)
pred <- predict(fit, dat_test, type = "response") > 0.5
table(pred, x[-train, ]$highcrim)
mean(pred != x$highcrim[-train]) * 100
summary(fit)
```
