# Linear Model Selection and Regularization

## Conceptual

### Question 1

> We perform best subset, forward stepwise, and backward stepwise selection on
> a single data set. For each approach, we obtain $p + 1$ models, containing
> $0, 1, 2, ..., p$ predictors. Explain your answers:
>
> a. Which of the three models with $k$ predictors has the smallest *training*
>    RSS?

Best subset considers the most models (all possible combinations of $p$ 
predictors are considered), therefore this will give the smallest training RSS
(it will at least consider all possibilities covered by forward and backward
stepwise selection). However, all three approaches are expected to give similar
if not identical results in practice.

> b. Which of the three models with $k$ predictors has the smallest *test* RSS?

We cannot tell which model will perform best on the test RSS. The answer will
depend on the tradeoff between fitting to the data and overfitting.

> c. True or False:
>    i. The predictors in the $k$-variable model identified by forward stepwise
>       are a subset of the predictors in the ($k+1$)-variable model identified
>       by forward stepwise selection.

True. Forward stepwise selection retains all features identified in previous
models as $k$ is increased.

>    ii. The predictors in the $k$-variable model identified by backward stepwise
>       are a subset of the predictors in the $(k+1)$-variable model identified
>       by backward stepwise selection.

True. Backward stepwise selection removes features one by one as $k$ is 
decreased.

>    iii. The predictors in the $k$-variable model identified by backward
>       stepwise are a subset of the predictors in the $(k+1)$-variable model
>       identified by forward stepwise selection.

False. Forward and backward stepwise selection can identify different 
combinations of variables due to differing algorithms.

>    iv. The predictors in the $k$-variable model identified by forward stepwise
>       are a subset of the predictors in the $(k+1)$-variable model identified
>       by backward stepwise selection.

False. Forward and backward stepwise selection can identify different 
combinations of variables due to differing algorithms.

>    v. The predictors in the $k$-variable model identified by best subset are a
>       subset of the predictors in the $(k+1)$-variable model identified by best
>       subset selection.

False. Best subset selection can identify different combinations of variables
for each $k$ by considering all possible models.

### Question 2

> For parts (a) through (c), indicate which of i. through iv. is correct.
> Justify your answer.
>
> a. The lasso, relative to least squares, is:
>    i. More flexible and hence will give improved prediction accuracy when its
>      increase in bias is less than its decrease in variance.
>    ii. More flexible and hence will give improved prediction accuracy when its
>      increase in variance is less than its decrease in bias.
>    iii. Less flexible and hence will give improved prediction accuracy when its
>      increase in bias is less than its decrease in variance.
>    iv. Less flexible and hence will give improved prediction accuracy when
>      its increase in variance is less than its decrease in bias.

iii. By using shrinkage, lasso can reduce the number of predictors so is less
flexible. As a result, it will lead to an increase in bias by approximating the
true relationship. We hope that this increase is small but that we dramatically
reduce variance (i.e. the difference we would see in the model fit between
different sets of training data).

> b. Repeat (a) for ridge regression relative to least squares.

iii. The same is true of ridge regression---shrinkage results in a less
flexible model and can reduce variance.

> c. Repeat (a) for non-linear methods relative to least squares.

ii. Non-linear methods can be more flexible. They can perform better as long as 
they don't substantially increase variance.

### Question 3

> Suppose we estimate the regression coefficients in a linear regression model
> by minimizing:
>
> $$
> \sum_{i=1}^n\left(y_i - \beta_0 - \sum_{j=1}^p\beta_jx_{ij}\right)^2
>   \textrm{subject to} \sum_{j=1}^p|\beta_j| \le s
> $$
>
> for a particular value of $s$. For parts (a) through (e), indicate
> which of i. through v. is correct. Justify your answer.
>
> a. As we increase $s$ from 0, the training RSS will:
>    i. Increase initially, and then eventually start decreasing in an
>      inverted U shape.
>    ii. Decrease initially, and then eventually start increasing in a U shape.
>    iii. Steadily increase.
>    iv. Steadily decrease.
>    v. Remain constant.

iv. As $s$ increases, the model becomes more flexible (the sum of absolute 
coefficients can be higher). With more flexible models, training RSS will
always decrease.

> b. Repeat (a) for test RSS.

ii. With more flexible models, test RSS will decrease (as the fit improves) and
will then increase due to overfitting (high variance).

> c. Repeat (a) for variance.

iii. As $s$ increases, the model becomes more flexible so variance will
increase.

> d. Repeat (a) for (squared) bias.

iv. As $s$ increases, the model becomes more flexible so bias will decrease.

> e. Repeat (a) for the irreducible error.

v. The irreducible error is unchanged.

### Question 4

> Suppose we estimate the regression coefficients in a linear regression model
> by minimizing
>
> $$
> \sum_{i=1}^n \left(y_i - \beta_0 - \sum_{j=1}^p\beta_jx_{ij}\right)^2 +
>   \lambda\sum_{j=1}^p\beta_j^2
> $$
>
> for a particular value of $\lambda$. For parts (a) through (e), indicate which
> of i. through v. is correct. Justify your answer.
>
> a. As we increase $\lambda$ from 0, the training RSS will:
>    i. Increase initially, and then eventually start decreasing in an
>    inverted U shape.
>    ii. Decrease initially, and then eventually start increasing in a U shape.
>    iii. Steadily increase.
>    iv. Steadily decrease.
>    v. Remain constant.

iii. As $\lambda$ is increased, more weight is placed on the sum of squared
coefficients and so the model becomes less flexible. As a result, training RSS
must increase.

> b. Repeat (a) for test RSS.

ii. As $\lambda$ increases, flexibility decreases so test RSS will decrease 
(variance decreases) but will then increase (as bias increases).

> c. Repeat (a) for variance.

iv. Steadily decrease.

> d. Repeat (a) for (squared) bias.

iii. Steadily increase.

> e. Repeat (a) for the irreducible error.

v. The irreducible error is unchanged.

### Question 5

> It is well-known that ridge regression tends to give similar coefficient
> values to correlated variables, whereas the lasso may give quite different
> coefficient values to correlated variables. We will now explore this property
> in a very simple setting.
>
> Suppose that $n = 2, p = 2, x_{11} = x_{12}, x_{21} = x_{22}$. Furthermore,
> suppose that $y_1 + y_2 =0$ and $x_{11} + x_{21} = 0$ and
> $x_{12} + x_{22} = 0$, so that the estimate for the intercept in a least
> squares, ridge regression, or lasso model is zero: $\hat{\beta}_0 = 0$.
>
> a. Write out the ridge regression optimization problem in this setting.

We are trying to minimize:

$$
\sum_{i=1}^n \left(y_i - \beta_0 - \sum_{j=1}^p\beta_jx_{ij}\right)^2 +
  \lambda\sum_{j=1}^p\beta_j^2
$$

We can ignore $\beta_0$ and can expand the sums since there's only two terms.
Additionally, we can define $x_1 = x_{11} = x_{12}$ and 
$x_2 = x_{21} = x_{22}$. We then need to minimize

\begin{align}
f = & (y_1 - \beta_1x_1 - \beta_2x_1)^2 + 
     (y_2 - \beta_1x_2 - \beta_2x_2)^2 + 
     \lambda\beta_1^2 + \lambda\beta_2^2 \\

f = & y_1^2 - 2y_1\beta_1x_1 - 2y_1\beta_2x_1 + \beta_1^2x_1^2 + 2\beta_1\beta_2x_1^2 + \beta_2^2x_1^2 + \\
    & y_2^2 - 2y_2\beta_1x_2 - 2y_2\beta_2x_2 + \beta_1^2x_2^2 + 2\beta_1\beta_2x_2^2 + \beta_2^2x_2^2 + \\
    & \lambda\beta_1^2 + \lambda\beta_2^2 \\
\end{align}

> b. Argue that in this setting, the ridge coefficient estimates satisfy
>    $\hat{\beta}_1 = \hat{\beta}_2$

We can find when the above is minimized with respect to each of $\beta_1$ and
$\beta_2$ by partial differentiation.

$$
\frac{\partial}{\partial{\beta_1}} = 
  - 2y_1x_1 + 2\beta_1x_1^2 + 2\beta_2x_1^2
  - 2y_2x_2 + 2\beta_1x_2^2 + 2\beta_2x_2^2
  + 2\lambda\beta_1
$$

$$
\frac{\partial}{\partial{\beta_2}} = 
  - 2y_1x_1 + 2\beta_1x_1^2 + 2\beta_2x_1^2
  - 2y_2x_2 + 2\beta_1x_2^2 + 2\beta_2x_2^2
  + 2\lambda\beta_2
$$

A minimum can be found when these are set to 0.

$$
\lambda\beta_1 = y_1x_1 + y_2x_2 - \beta_1x_1^2 - \beta_2x_1^2 - \beta_1x_2^2 - \beta_2x_2^2 \\
\lambda\beta_2 = y_1x_1 + y_2x_2 - \beta_1x_1^2 - \beta_2x_1^2 - \beta_1x_2^2 - \beta_2x_2^2 
$$

Therefore $\lambda\beta_1 = \lambda\beta_2$ and $\beta_1 = \beta_2$, thus
there is only one solution, that is when the coefficients are the same.

> c. Write out the lasso optimization problem in this setting.

We are trying to minimize:

$$
\sum_{i=1}^n \left(y_i - \beta_0 - \sum_{j=1}^p\beta_jx_{ij}\right)^2 +
  \lambda\sum_{j=1}^p |\beta_j|
$$

As above (and defining $x_1 = x_{11} = x_{12}$ and $x_2 = x_{21} = x_{22}$) we simplify to

$$
(y_1 - \beta_1x_1 - \beta_2x_1)^2 + 
  (y_2 - \beta_1x_2 - \beta_2x_2)^2 + 
  \lambda|\beta_1| + \lambda|\beta_2|
$$

> d. Argue that in this setting, the lasso coefficients $\hat{\beta}_1$ and
>    $\hat{\beta}_2$ are not unique---in other words, there are many possible
>    solutions to the optimization problem in (c). Describe these solutions.

We will consider the alternate form of the lasso optimization problem

$$
(y_1 - \hat{\beta_1}x_1 - \hat{\beta_2}x_1)^2 + (y_2 - \hat{\beta_1}x_2 - \hat{\beta_2}x_2)^2 \quad \text{subject to} \quad |\hat{\beta_1}| + |\hat{\beta_2}| \le s
$$

Since $x_1 + x_2 = 0$ and $y_1 + y_2 = 0$, this is equivalent to minimising
$2(y_1 - (\hat{\beta_1} + \hat{\beta_2})x_1)^2$
which has a solution when $\hat{\beta_1} + \hat{\beta_2} = y_1/x_1$.
Geometrically, this is a $45^\circ$ backwards sloping line in the 
($\hat{\beta_1}$, $\hat{\beta_2}$) plane.

The constraints $|\hat{\beta_1}| + |\hat{\beta_2}| \le s$ specify a diamond 
shape in the same place, also with lines that are at $45^\circ$ centered at the
origin and which intersect the axes at a distance $s$ from the origin. 

Thus, points along two edges of the diamond
($\hat{\beta_1} + \hat{\beta_2} = s$ and $\hat{\beta_1} + \hat{\beta_2} = -s$) 
become solutions to the lasso optimization problem.

### Question 6

> We will now explore (6.12) and (6.13) further.
>
> a. Consider (6.12) with $p = 1$. For some choice of $y_1$ and $\lambda > 0$,
>    plot (6.12) as a function of $\beta_1$. Your plot should confirm that
>    (6.12) is solved by (6.14).

Equation 6.12 is:

$$
\sum_{j=1}^p(y_j - \beta_j)^2 + \lambda\sum_{j=1}^p\beta_j^2
$$

Equation 6.14 is:

$$
\hat{\beta}_j^R = y_j/(1 + \lambda)
$$

where $\hat{\beta}_j^R$ is the ridge regression estimate.

```{r}
lambda <- 0.7
y <- 1.4
fn <- function(beta) {
  (y - beta)^2 + lambda * beta^2
}
plot(seq(0, 2, 0.01), fn(seq(0, 2, 0.01)), type = "l", xlab = "beta", ylab = "6.12")
abline(v = y / (1 + lambda), lty = 2)
```

> b. Consider (6.13) with $p = 1$. For some choice of $y_1$ and $\lambda > 0$,
>    plot (6.13) as a function of $\beta_1$. Your plot should confirm that
>    (6.13) is solved by (6.15).

Equation 6.13 is:

$$
\sum_{j=1}^p(y_j - \beta_j)^2 + \lambda\sum_{j=1}^p|\beta_j|
$$

Equation 6.15 is:

$$
\hat{\beta}_j^L = \begin{cases}
  y_j - \lambda/2 &\mbox{if } y_j > \lambda/2; \\
  y_j + \lambda/2 &\mbox{if } y_j < -\lambda/2; \\
  0               &\mbox{if } |y_j| \le \lambda/2;
\end{cases}
$$

For $\lambda = 0.7$ and $y = 1.4$, the top case applies.

```{r}
lambda <- 0.7
y <- 1.4
fn <- function(beta) {
  (y - beta)^2 + lambda * abs(beta)
}
plot(seq(0, 2, 0.01), fn(seq(0, 2, 0.01)), type = "l", xlab = "beta", ylab = "6.12")
abline(v = y - lambda / 2, lty = 2)
```

### Question 7

> We will now derive the Bayesian connection to the lasso and ridge regression
> discussed in Section 6.2.2.
>
> a. Suppose that $y_i = \beta_0 + \sum_{j=1}^p x_{ij}\beta_j + \epsilon_i$
>    where $\epsilon_1, ..., \epsilon_n$ are independent and identically
>    distributed from a $N(0, \sigma^2)$ distribution. Write out the likelihood
>    for the data.

\begin{align*}
\mathcal{L} 
  &= \prod_i^n \mathcal{N}(0, \sigma^2) \\
  &= \prod_i^n \frac{1}{\sqrt{2\pi\sigma}}\exp\left(-\frac{\epsilon_i^2}{2\sigma^2}\right) \\
  &= \left(\frac{1}{\sqrt{2\pi\sigma}}\right)^n \exp\left(-\frac{1}{2\sigma^2} \sum_i^n \epsilon_i^2\right)
\end{align*}

> b. Assume the following prior for $\beta$: $\beta_1, ..., \beta_p$ are
>    independent and identically distributed according to a double-exponential
>    distribution with mean 0 and common scale parameter b: i.e.
>    $p(\beta) = \frac{1}{2b}\exp(-|\beta|/b)$. Write out the posterior for
>    $\beta$ in this setting.

The posterior can be calculated by multiplying the prior and likelihood
(up to a proportionality constant).

\begin{align*}
p(\beta|X,Y) 
  &\propto \left(\frac{1}{\sqrt{2\pi\sigma}}\right)^n \exp\left(-\frac{1}{2\sigma^2} \sum_i^n \epsilon_i^2\right) \prod_j^p\frac{1}{2b}\exp\left(-\frac{|\beta_j|}{b}\right)  \\
  &\propto \frac{1}{2b} \left(\frac{1}{\sqrt{2\pi\sigma}}\right)^n \exp\left(-\frac{1}{2\sigma^2} \sum_i^n \epsilon_i^2 -\sum_j^p\frac{|\beta_j|}{b}\right)
\end{align*}

> c. Argue that the lasso estimate is the _mode_ for $\beta$ under this
>    posterior distribution.

Let us find the maximum of the posterior distribution (the mode). Maximizing
the posterior probability is equivalent to maximizing its log which is:

$$
\log(p(\beta|X,Y)) \propto  \log\left[ \frac{1}{2b} \left(\frac{1}{\sqrt{2\pi\sigma}}\right)^n \right ] - \left(\frac{1}{2\sigma^2} \sum_i^n \epsilon_i^2 + \sum_j^p\frac{|\beta_j|}{b}\right)
$$

Since, the first term is independent of $\beta$, our solution will be when
we minimize the second term.

\begin{align*}
\DeclareMathOperator*{\argmin}{arg\,min} % Jan Hlavacek
\argmin_\beta \left(\frac{1}{2\sigma^2} \sum_i^n \epsilon_i^2 + \sum_j^p\frac{|\beta|}{b}\right)
&= \argmin_\beta \left(\frac{1}{2\sigma^2} \right ) \left( \sum_i^n \epsilon_i^2 +\frac{2\sigma^2}{b}\sum_j^p|\beta_j|\right) \\
&= \argmin_\beta \left( \sum_i^n \epsilon_i^2 +\frac{2\sigma^2}{b}\sum_j^p|\beta_j|\right)
\end{align*}

Note, that $RSS = \sum_i^n \epsilon_i^2$ and if we set $\lambda =
\frac{2\sigma^2}{b}$, the mode corresponds to lasso optimization.
$$
\argmin_\beta RSS + \lambda\sum_j^p|\beta_j|
$$

> d. Now assume the following prior for $\beta$: $\beta_1, ..., \beta_p$ are
>    independent and identically distributed according to a normal distribution
>    with mean zero and variance $c$. Write out the posterior for $\beta$ in
>    this setting.

The posterior is now:

\begin{align*}
p(\beta|X,Y) 
  &\propto \left(\frac{1}{\sqrt{2\pi\sigma}}\right)^n \exp\left(-\frac{1}{2\sigma^2} \sum_i^n \epsilon_i^2\right) \prod_j^p\frac{1}{\sqrt{2\pi c}}\exp\left(-\frac{\beta_j^2}{2c}\right)  \\
  &\propto 
   \left(\frac{1}{\sqrt{2\pi\sigma}}\right)^n 
   \left(\frac{1}{\sqrt{2\pi c}}\right)^p
\exp\left(-\frac{1}{2\sigma^2} \sum_i^n \epsilon_i^2 - \frac{1}{2c}\sum_j^p\beta_j^2\right)
\end{align*}

> e. Argue that the ridge regression estimate is both the _mode_ and the _mean_
>    for $\beta$ under this posterior distribution.

To show that the ridge estimate is the mode we can again find the maximum by
maximizing the log of the posterior. The log is 

$$
\log{p(\beta|X,Y)}
  \propto 
   \log{\left[\left(\frac{1}{\sqrt{2\pi\sigma}}\right)^n \left(\frac{1}{\sqrt{2\pi c}}\right)^p \right ]}
- \left(\frac{1}{2\sigma^2} \sum_i^n \epsilon_i^2 + \frac{1}{2c}\sum_j^p\beta_j^2 \right)
$$

We can maximize (wrt $\beta$) by ignoring the first term and minimizing the
second term. i.e. we minimize:

$$
\argmin_\beta \left( \frac{1}{2\sigma^2} \sum_i^n \epsilon_i^2 + \frac{1}{2c}\sum_j^p\beta_j^2 \right)\\
= \argmin_\beta \left( \frac{1}{2\sigma^2} \left( \sum_i^n \epsilon_i^2 + \frac{\sigma^2}{c}\sum_j^p\beta_j^2 \right) \right)
$$

As above, if $RSS = \sum_i^n \epsilon_i^2$ and if we set $\lambda =
\frac{\sigma^2}{c}$, we can see that the mode corresponds to ridge optimization.

## Applied

### Question 8

> In this exercise, we will generate simulated data, and will then use this
> data to perform best subset selection.
>
> a. Use the `rnorm()` function to generate a predictor $X$ of length $n = 100$,
>    as well as a noise vector $\epsilon$ of length $n = 100$.

```{r, message = FALSE, warning = FALSE}
library(ISLR2)
library(glmnet)
library(leaps)
library(pls)
```

```{r}
set.seed(42)
x <- rnorm(100)
ep <- rnorm(100)
```

> b. Generate a response vector $Y$ of length $n = 100$ according to the model
>    $$Y = \beta_0 + \beta_1X + \beta_2X^2 + \beta_3X^3 + \epsilon,$$
>    where $\beta_0, \beta_1, \beta_2,$ and $\beta_3$ are constants of your
>    choice.

```{r}
y <- 2 + 3 * x - 2 * x^2 + 0.5 * x^3 + ep
```

> c. Use the `regsubsets()` function to perform best subset selection in order
>    to choose the best model containing the predictors $X, X^2, ..., X^{10}$.
>    What is the best model obtained according to $C_p$, BIC, and adjusted
>    $R^2$? Show some plots to provide evidence for your answer, and report the
>    coefficients of the best model obtained. Note you will need to use the
>    `data.frame()` function to create a single data set containing both $X$ and
>    $Y$.

```{r}
dat <- data.frame(x, y)
summary(regsubsets(y ~ poly(x, 10, raw = TRUE), data = dat))
```

> d. Repeat (c), using forward stepwise selection and also using backwards
>    stepwise selection. How does your answer compare to the results in (c)?

```{r}
summary(regsubsets(y ~ poly(x, 10, raw = TRUE), data = dat, method = "forward"))
summary(regsubsets(y ~ poly(x, 10, raw = TRUE), data = dat, method = "backward"))
```

> e. Now fit a lasso model to the simulated data, again using
>    $X, X^2, ..., X^{10}$ as predictors. Use cross-validation to select the
>    optimal value of $\lambda$. Create plots of the cross-validation error as a
>    function of $\lambda$. Report the resulting coefficient estimates, and
>    discuss the results obtained.

```{r}
res <- cv.glmnet(poly(dat$x, 10, raw = TRUE), dat$y, alpha = 1)
(best <- res$lambda.min)
plot(res)
out <- glmnet(poly(dat$x, 10, raw = TRUE), dat$y, alpha = 1, lambda = res$lambda.min)
predict(out, type = "coefficients", s = best)
```

When fitting lasso, the model that minimizes MSE uses three predictors (as per
the simulation). The coefficients estimated (2.9, -1.9 and 0.5) are similar to 
those used in the simulation.

> f. Now generate a response vector $Y$ according to the model
>    $$Y = \beta_0 + \beta_7X^7 + \epsilon,$$ and perform best subset selection
>    and the lasso. Discuss the results obtained.

```{r}
dat$y <- 2 - 2 * x^2 + 0.2 * x^7 + ep
summary(regsubsets(y ~ poly(x, 10, raw = TRUE), data = dat))

res <- cv.glmnet(poly(dat$x, 10, raw = TRUE), dat$y, alpha = 1)
(best <- res$lambda.min)
plot(res)
out <- glmnet(poly(dat$x, 10, raw = TRUE), dat$y, alpha = 1, lambda = best)
predict(out, type = "coefficients", s = best)
```

When fitting lasso, the model does not perfectly replicate the simulation 
(coefficients are retained for powers of $x$ that were not simulated).

### Question 9

> In this exercise, we will predict the number of applications received using
> the other variables in the `College` data set.
>
> a. Split the data set into a training set and a test set.

```{r}
set.seed(42)
train <- sample(nrow(College), nrow(College) * 2 / 3)
test <- setdiff(seq_len(nrow(College)), train)
mse <- list()
```

> b. Fit a linear model using least squares on the training set, and report the
>    test error obtained.

```{r}
fit <- lm(Apps ~ ., data = College[train, ])
(mse$lm <- mean((predict(fit, College[test, ]) - College$Apps[test])^2))
```

> c. Fit a ridge regression model on the training set, with $\lambda$ chosen by
>    cross-validation. Report the test error obtained.

```{r}
mm <- model.matrix(Apps ~ ., data = College[train, ])
fit2 <- cv.glmnet(mm, College$Apps[train], alpha = 0)
p <- predict(fit2, model.matrix(Apps ~ ., data = College[test, ]), s = fit2$lambda.min)
(mse$ridge <- mean((p - College$Apps[test])^2))
```

> d. Fit a lasso model on the training set, with $\lambda$ chosen by cross-
>    validation. Report the test error obtained, along with the number of
>    non-zero coefficient estimates.

```{r}
mm <- model.matrix(Apps ~ ., data = College[train, ])
fit3 <- cv.glmnet(mm, College$Apps[train], alpha = 1)
p <- predict(fit3, model.matrix(Apps ~ ., data = College[test, ]), s = fit3$lambda.min)
(mse$lasso <- mean((p - College$Apps[test])^2))
```

> e. Fit a PCR model on the training set, with $M$ chosen by cross-validation.
>    Report the test error obtained, along with the value of $M$ selected by
>    cross-validation.

```{r}
fit4 <- pcr(Apps ~ ., data = College[train, ], scale = TRUE, validation = "CV")
validationplot(fit4, val.type = "MSEP")
p <- predict(fit4, College[test, ], ncomp = 17)
(mse$pcr <- mean((p - College$Apps[test])^2))
```

> f. Fit a PLS model on the training set, with $M$ chosen by cross-validation.
>    Report the test error obtained, along with the value of $M$ selected by
>    cross-validation.

```{r}
fit5 <- plsr(Apps ~ ., data = College[train, ], scale = TRUE, validation = "CV")
validationplot(fit5, val.type = "MSEP")
p <- predict(fit5, College[test, ], ncomp = 12)
(mse$pls <- mean((p - College$Apps[test])^2))
```

> g. Comment on the results obtained. How accurately can we predict the number
>    of college applications received? Is there much difference among the test
>    errors resulting from these five approaches?

```{r}
barplot(unlist(mse), ylab = "Test MSE", horiz = TRUE)
```

Ridge and lasso give the lowest test errors but the lowest is generated by
the ridge regression model (in this specific case with this specific seed).


### Question 10

> We have seen that as the number of features used in a model increases, the
> training error will necessarily decrease, but the test error may not. We will
> now explore this in a simulated data set.
>
> a. Generate a data set with $p = 20$ features, $n = 1,000$ observations, and
>    an associated quantitative response vector generated according to the model
>    $Y =X\beta + \epsilon$, where $\beta$ has some elements that are exactly
>    equal to zero.

```{r}
set.seed(42)
dat <- matrix(rnorm(1000 * 20), nrow = 1000)
colnames(dat) <- paste0("b", 1:20)
beta <- rep(0, 20)
beta[1:4] <- c(5, 4, 2, 7)
y <- colSums((t(dat) * beta)) + rnorm(1000)
dat <- data.frame(dat)
dat$y <- y
```

> b. Split your data set into a training set containing 100 observations and a
>    test set containing 900 observations.

```{r}
train <- dat[1:100, ]
test <- dat[101:1000, ]
```

> c. Perform best subset selection on the training set, and plot the training
>    set MSE associated with the best model of each size.

```{r}
fit <- regsubsets(y ~ ., data = train, nvmax = 20)
summary(fit)
plot(summary(fit)$rss / 100, ylab = "MSE", type = "o")
```

> d. Plot the test set MSE associated with the best model of each size.

```{r}
predict.regsubsets <- function(object, newdata, id, ...) {
  form <- as.formula(object$call[[2]])
  mat <- model.matrix(form, newdata)
  coefi <- coef(object, id = id)
  xvars <- names(coefi)
  mat[, xvars] %*% coefi
}
mse <- sapply(1:20, function(i) mean((test$y - predict(fit, test, i))^2))
plot(mse, ylab = "MSE", type = "o", pch = 19)
```

> e. For which model size does the test set MSE take on its minimum value?
>    Comment on your results. If it takes on its minimum value for a model
>    containing only an intercept or a model containing all of the features,
>    then play around with the way that you are generating the data in (a) until
>    you come up with a scenario in which the test set MSE is minimized for an
>    intermediate model size.

```{r}
which.min(mse)
```

The min test MSE is found when model size is 4. This corresponds to the 
simulated data which has four non-zero coefficients.

```{r}
set.seed(42)
dat <- matrix(rnorm(1000 * 20), nrow = 1000)
colnames(dat) <- paste0("b", 1:20)
beta <- rep(0, 20)
beta[1:9] <- c(5, 4, 2, 7, 0.01, 0.001, 0.05, 0.1, 0.5)
y <- colSums((t(dat) * beta)) + rnorm(1000)
dat <- data.frame(dat)
dat$y <- y
train <- dat[1:100, ]
test <- dat[101:1000, ]

fit <- regsubsets(y ~ ., data = train, nvmax = 20)
summary(fit)
mse <- sapply(1:20, function(i) mean((test$y - predict(fit, test, i))^2))
plot(mse, ylab = "MSE", type = "o", pch = 19)
which.min(mse)
```

> f. How does the model at which the test set MSE is minimized compare to the
>    true model used to generate the data? Comment on the coefficient values.

The min test MSE is found when model size is 5 but there are 9 non-zero 
coefficients.

```{r}
coef(fit, id = 5)
```

The coefficient values are well estimated when high, but the smaller 
coefficients are dropped.

> g. Create a plot displaying $\sqrt{\sum_{j=1}^p (\beta_j - \hat{\beta}{}_j^r)^2}$
>    for a range of values of $r$, where $\hat{\beta}{}_j^r$ is the $j$th
>    coefficient estimate for the best model containing $r$ coefficients. Comment
>    on what you observe. How does this compare to the test MSE plot from (d)?

```{r}
names(beta) <- paste0("b", 1:20)
b <- data.frame(id = names(beta), b = beta)

out <- sapply(1:20, function(i) {
  c <- coef(fit, id = i)[-1]
  c <- data.frame(id = names(c), c = c)
  m <- merge(b, c)
  sqrt(sum((m$b - m$c)^2))
})
plot(out, ylab = "Mean squared coefficient error", type = "o", pch = 19)
```

The error of the coefficient estimates is minimized when model size is 5. This
corresponds to the point when test MSE was minimized.

### Question 11

> We will now try to predict per capita crime rate in the `Boston` data set.
>
> a. Try out some of the regression methods explored in this chapter, such as
>    best subset selection, the lasso, ridge regression, and PCR. Present and
>    discuss results for the approaches that you consider.

```{r}
set.seed(1)
train <- sample(nrow(Boston), nrow(Boston) * 2 / 3)
test <- setdiff(seq_len(nrow(Boston)), train)
hist(log(Boston$crim))
```

> b. Propose a model (or set of models) that seem to perform well on this data
>    set, and justify your answer. Make sure that you are evaluating model
>    performance using validation set error, cross-validation, or some other
>    reasonable alternative, as opposed to using training error.

We will try to fit models to `log(Boston$crim)` which is closer to a normal
distribution.

```{r}
fit <- lm(log(crim) ~ ., data = Boston[train, ])
mean((predict(fit, Boston[test, ]) - log(Boston$crim[test]))^2)

mm <- model.matrix(log(crim) ~ ., data = Boston[train, ])
fit2 <- cv.glmnet(mm, log(Boston$crim[train]), alpha = 0)
p <- predict(fit2, model.matrix(log(crim) ~ ., data = Boston[test, ]), s = fit2$lambda.min)
mean((p - log(Boston$crim[test]))^2)

mm <- model.matrix(log(crim) ~ ., data = Boston[train, ])
fit3 <- cv.glmnet(mm, log(Boston$crim[train]), alpha = 1)
p <- predict(fit3, model.matrix(log(crim) ~ ., data = Boston[test, ]), s = fit3$lambda.min)
mean((p - log(Boston$crim[test]))^2)

fit4 <- pcr(log(crim) ~ ., data = Boston[train, ], scale = TRUE, validation = "CV")
validationplot(fit4, val.type = "MSEP")
p <- predict(fit4, Boston[test, ], ncomp = 8)
mean((p - log(Boston$crim[test]))^2)

fit5 <- plsr(log(crim) ~ ., data = Boston[train, ], scale = TRUE, validation = "CV")
validationplot(fit5, val.type = "MSEP")
p <- predict(fit5, Boston[test, ], ncomp = 6)
mean((p - log(Boston$crim[test]))^2)
```

In this case lasso (`alpha = 1`) seems to perform very slightly better than
un-penalized regression. Some coefficients have been dropped:

```{r}
coef(fit3, s = fit3$lambda.min)
```

> c. Does your chosen model involve all of the features in the data set? Why or
>    why not?

Not all features are included due to the lasso penalization.
